DOpElib: Differential Equations and Optimization Environment; A Goal Oriented Software Library for Solving PDEs and Optimization Problems

### 2012-11-27 Paper_May_2012/main.tex
Aenderungen und Kommentare von cg ab 2.2 bis zum Ende

\subsection{Solving Nonstationary PDEs}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form $a(u)(\phi)$ and some
right hand side functional $l(\phi)$.
Let $I:=[0,T]$ be a time interval with end time point value $T$.
To find some $u$ in a weak form, we define
\[
\int_I a(u)(\phi)\, dt = \int_I l(\phi)\, dt \quad \forall\Phi\in X,
\]
with some suitable Bochner space $X$.  As model problem,
we extend the previous example and consider the nonstationary Navier-Stokes equations
in a domain $\Omega\in\mathbb{R}^d$. Find: 
$(v,p)\in X$ with
\begin{align*}
&\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt\\
&\quad + (v(0) - v^0, \phi(0))
= \int_I l(\phi) \, dt 
\end{align*}
for all $\Phi:= (\phi, \chi) \in X$ with
\[
X:= \bigl\{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d) , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R} \bigr\}.
\]
As for stationary problems, the goal is to compute
some target functional $J$ which is now possibly 
time-dependent. To organize the discretization 
process, we need to perform spatial and temporal 
discretization. This is again done via
an object that takes care of organizing the degrees of 
freedom, as in the case of stationary problems.
The only difference to the stationary case is that we need to 
add the temporal subdivision \texttt{times}. In the simplest 
case, when we have the same spatial triangulation 
at all times, the constructor could look as follows:
\begin{lstlisting}
  template<typename TIMES, typename TRIANGULATION, 
           typename FE>
   void SpaceTimeHandler(TIMES& times,
                         TRIANGULATION& triag,
                         FE& fe);
\end{lstlisting}

Now we may discretize the problem with some time-stepping scheme, which will 
remove the temporal derivative and replace it by some difference quotients. 
Since the exact way of doing this depends on the time-stepping scheme 
we need a new object that takes the user-given description of the PDE and 
generates the appropriate combinations of spatial integrals and scaling.
Thus a generic problem may look as follows:
\begin{lstlisting}
  template<typename PROBLEM>
      class TimeStepProblem(Problem & P);
\end{lstlisting}
At present we have prepared the following time-stepping schemes in \dope:
\begin{itemize}
\item Forward-(or Explicit)-Euler
\item Backward-(or Implicit)-Euler
\item Crank-Nicolson scheme
\item shifted Crank-Nicolson scheme (to deal with singular initial data)
\item fractional-step-$\theta$ scheme
\end{itemize}
After choosing the time-stepping scheme, the SpaceTimeHandler  
loops over all time-steps, where in each of the 
time-steps we can reuse the functionality we have provided to solve 
stationary PDEs. 
This means for instance, if we have implemented the 
cell-integrals in the weak formulation for the stationary case, 
then we can use the exact same 
implementation again for the solution of the nonstationary version of 
the equation. However, we need to provide an additional 
term to cope with the integral for the time derivative.

>>>>

\subsection{Solving Nonstationary PDEs}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form $a(u)(\phi)$ and some
right hand side functional $l$.\todocg{Warum hier l und nicht f?}
Let $I:=[0,T]$ be a time interval with end time point value $T$.
To find some $u$ in a weak form, we define
\[
\int_I a(u)(\phi)\, dt = \int_I l(\phi)\, dt \quad \forall\Phi\in X,
\]
with some suitable Bochner space $X$.  As model problem,
we extend the previous example and consider the nonstationary Navier-Stokes equations
in a domain $\Omega\in\mathbb{R}^d$. Find: 
$(v,p)\in X$ with
\begin{align*}
&\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt\\
&\quad + (v(0) - v^0, \phi(0))
= \int_I l(\phi) \, dt 
\end{align*}
for all $\Phi:= (\phi, \chi) \in X$ with
\[
X:= \set{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d) , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R}}.
\]
As for stationary problems, the goal is to compute
some target functional $J$ which is now possibly 
time-dependent. To organize the discretization 
process, we need to perform spatial and temporal 
discretization. This is again done via
an object that takes care of organizing the degrees of 
freedom, as in the case of stationary problems.
The only difference to the stationary case is that we need to 
add the temporal subdivision \texttt{times}. In the simplest 
case, when we have the same spatial triangulation 
at all times, the constructor could look as follows:
\begin{lstlisting}
  template<typename TIMES, typename TRIANGULATION, 
           typename FE>
   void SpaceTimeHandler(TIMES& times,
                         TRIANGULATION& triag,
                         FE& fe);
\end{lstlisting}

Now we may discretize the problem with some time-stepping scheme, which will 
remove the temporal derivative and replace it by some difference quotients. 
Since the exact way of doing this depends on the time-stepping scheme 
we need a new object that takes the user-given description of the PDE and 
generates the appropriate combinations of spatial integrals and scaling.
Thus a generic problem may look as follows:
\begin{lstlisting}
  template<typename PROBLEM>
      class TimeStepProblem(Problem & P);
\end{lstlisting}
At present we have prepared the following time-stepping schemes in \dope:
\begin{itemize}
\item Forward-(or Explicit)-Euler
\item Backward-(or Implicit)-Euler
\item Crank-Nicolson scheme
\item Shifted Crank-Nicolson scheme (to deal with singular initial data)
\item Fractional-Step-$\theta$ scheme
\end{itemize}
After choosing the time-stepping scheme, the \texttt{SpaceTimeHandler}  
loops over all time-steps, where in each of the 
time-steps we can reuse the functionality we have provided to solve 
stationary PDEs. 
This means for instance, if we have implemented the 
cell-integrals in the weak formulation for the stationary case, 
then we can use the exact same 
implementation again for the solution of the nonstationary version of 
the equation. However, we need to provide an additional 
term to cope with the integral for the time derivative.
